2H qSIP Calculations

### 2021-10-21 qSIP_calculations.Rmd
full data pipeline, plots

```{r, message=FALSE, warning=FALSE}
library(tidyverse) # general data wrangling and plotting
library(isoreader) # reading the raw data files
library(isoprocessor) # processing the data
library(isotopia) # isotope utilities
library(readxl) # reading excel docs
library(ggsci) # some nice color palettes
library(ggborderline) # nice looking ggplot lines
library(ggrepel)
library(latex2exp)
library(ggdist)

@@

## Generate Isoverse Peak Map
```{r}
# Generate an isoverse peak map
peaks_mapped <- sample_peak_table %>%
  iso_map_peaks(peak_map, map_id = Analysis) %>%
  filter(!is.na(compound))

@@

```{r}

# Add Some Metadata

peaks_mapped <- peaks_mapped %>% 
  group_by(sample_id) %>% 
  mutate(inc_time_d = as.numeric(substr(sample_id, 4,4)),
         inc_time_d_str = paste(inc_time_d, "Days"),
         f_label = 0.005,
         t_series_id = paste0(substr(sample_id, 1,1), 
                              substr(sample_id, 5,5)))

inc_time_zero <- peaks_mapped %>% 
  filter(inc_time_d == 0) %>% 
  select(sample_id, compound, at2H, t_series_id) %>% 
  mutate(f_start = at2H) %>% 
  select(-at2H)

## Trying a simpler example
c_inc_time_zero <- inc_time_zero %>% 
  filter(str_detect(t_series_id, "c"))

c_peaks_mapped <- peaks_mapped %>% 
  filter(str_detect(t_series_id, "c")) %>% 
  full_join(c_inc_time_zero, by = c("t_series_id", "compound"))

@@

```{r}
zero_peaks <- peaks_mapped %>%
  filter(inc_time_d_str == "0 Days") %>%
  group_by(t_series_id, compound) %>%
  mutate(n_analyical_reps = n()) %>%
  ungroup() %>%
  # exclude ambiguous peaks and those with multiple data points
  # this needs to be fixed
  filter(!is_ambiguous, n_analyical_reps == 1)

peaks_mapped_with_zeros <- 
  peaks_mapped %>%
  left_join(
    zero_peaks %>%
      select(
        compound, t_series_id, zero_sample_id = sample_id, area2_zero = area2, d2H_zero = d2H
      ),
    by = c("t_series_id", "compound")
  ) 
stopifnot(nrow(peaks_mapped_with_zeros) == nrow(peaks_mapped))

@@

```{r}
#peak maps
peaks_mapped %>%
  ungroup() %>%
  select(Analysis, compound, rt) %>%
  mutate(rt = round(rt, digits=0)) %>%
  pivot_wider(
    names_from = Analysis, 
    names_prefix = "rt:",
    values_from = rt, 
    id_cols = compound, 
    values_fn = function(x) paste(x, collapse = "; ")
  ) %>%
  openxlsx::write.xlsx(file.path("data", "IRMS", "peak_maps_resolve.xlsx"))
```

```{r}
peaks_mapped <- peaks_mapped %>%
  group_by(sample_id) %>%
  mutate(f_start = case_when(
           inc_time_d == 0 ~ at2H,
           inc_time_d != 0 ~ inc_time_zero$at2H
           ),
         terrain = case_when(
           substr(sample_id, 1, 1) == "t" ~ "Niwot Ridge Tundra",
           substr(sample_id, 1 ,1) == "c" ~ "Gordon Gulch Conifer Forest",
           substr(sample_id, 1 ,1) == "m" ~ "Gordon Gulch Meadow",
           substr(sample_id, 1 ,1) == "g" ~ "Marshall Mesa Grassland"))

peaks_mapped <- peaks_mapped %>% 
  mutate(rel_amount = area2 / sum(area2[compound != "IS"]))
```

## Remove standards from gencalcs
```{r}
# Don't want our standard compounds to be used for generation time calculations! x_x
peaks_mapped_active <- peaks_mapped %>% 
  filter(str_detect(compound, "STD", negate = TRUE))
```

@@

```{r}
gen_calc_data <- peaks_mapped_active %>% 
  select(c(Analysis, compound, d2H, at2H, sample_id, inc_time_d, inc_time_d_str, terrain, f_start, rel_amount)) %>% 
  mutate(
    at2H_label = 0.005,
    f_label = 0.5, # label is 0.5 at%
    a = 0.5, # assimilation efficiency
    t0 = 0, # start time
    d_t = inc_time_d - t0, # change in time
    f_t = at2H, # restate 2F at time t
    u_d = gencalc(a = 0.5, # Assimilation efficiency
                  FT = f_t, # 2F at time t (at%)
                  F0 = f_start, # 2F at t0 (at%)
                  FL = f_label, # 2F of label (at%)
                  t = d_t), # change in time
         # Returns turnover (µ) in days^-1
         gen_d = log(2) / u_d # calculate generation time
  )
```

```{r}
peaks_mapped_active %>% ggplot(
  aes(x = inc_time_d, y = at2H, color = compound, group = compound)) +
  geom_point() +
  geom_borderline() +
  scale_color_viridis_d() +
  theme_minimal()
```

@@

```{r "shaded chromatogram", fig.width=8, fig.height=8}
p_peaks_mapped_active <- peaks_mapped_active %>% ggplot(
  aes(x = rt, y = amp2, fill = as.numeric(at2H), label = compound)) +
  geom_col(aes(width = rt_end - rt_start)) +
  scale_fill_viridis_c(option = "magma") +
  facet_wrap(~inc_time_d_str, ncol = 1) +
  geom_text_repel(size = 2, box.padding = 0.5, nudge_y = 150) +
  theme_classic() +
  coord_cartesian(ylim = c(0, 1300)) +
  labs(fill = TeX("$^2F$"),
         x = "Retention Time (s)",
         y = "Amplitude (mV)") +
  theme(strip.background = element_blank())

@@

```{r}
p_gen_time_dist <- gen_calc_data %>% filter(inc_time_d != 0) %>% 
  ggplot(
  aes(x = gen_d, 
      y = inc_time_d_str, 
      color = inc_time_d_str, 
      fill = inc_time_d_str)) +
  stat_slab(height = 0.3, alpha = 0.7) +
  stat_pointinterval(color = "black") +
  geom_point(
    color = "black",
    shape = 124,
    size = 3,
    alpha = 1,
    position = position_nudge(y = .03)) +
  geom_text_repel(aes(label = compound), 
                  color = "black", 
                  nudge_y = -0.3, 
                  size = 3) +
  scale_x_log10() +
  annotation_logticks(sides = "b") +
  scale_fill_npg() +
  scale_color_npg() +
  theme_classic() +
  labs(x = "Generation time (days)",
       y = "Incubation time",
       title = "Generation times of soil bacteria",
       subtitle = "Lipid-specific generation times of bacteria collected from Alpine Tundra",
       caption = "Soils collected from Niwot Ridge LTER") +
  theme(legend.position = "none",
        axis.ticks.y = element_blank())
p_gen_time_dist

@@

>>>>

```{r, message=FALSE, warning=FALSE}
library(tidyverse) # general data wrangling and plotting
library(isoreader) # reading the raw data files
library(isoprocessor) # processing the data
library(isotopia) # isotope utilities
library(readxl) # reading excel docs
library(ggsci) # some nice color palettes
library(ggborderline) # nice looking ggplot lines
library(ggrepel)
library(latex2exp)
library(ggdist)
library(ggsignif)

@@

## Generate Isoverse Peak Map
```{r}
# Generate an isoverse peak map
peaks_mapped <- sample_peak_table %>%
  iso_map_peaks(peak_map, map_id = Analysis) %>%
  filter(!is.na(compound))
```

@@

## Export summary of unresolved peaks
```{r}
# Spit out peaks that need to be resolved
peaks_mapped %>%
  ungroup() %>%
  select(Analysis, compound, rt) %>%
  mutate(rt = round(rt, digits=0)) %>%
  pivot_wider(
    names_from = Analysis, 
    names_prefix = "rt:",
    values_from = rt, 
    id_cols = compound, 
    values_fn = function(x) paste(x, collapse = "; ")
  ) %>%
  openxlsx::write.xlsx(file.path("data", "IRMS", "peak_maps_resolve.xlsx"))
```

## Assign incubation params
```{r}

# Add Some Parameters
peaks_mapped <- peaks_mapped %>% 
  group_by(sample_id) %>% 
  mutate(inc_time_d = as.numeric(substr(sample_id, 4,4)),
         inc_time_d_str = paste(inc_time_d, "Days"),
         f_label = 0.005,
         t_series_id = paste0(substr(sample_id, 1,1), 
                              substr(sample_id, 5,5)),
         soil_id = substr(t_series_id, 1,1),
         replicate_id = substr(t_series_id, 2,2),
         terrain = case_when(
           substr(sample_id, 1, 1) == "t" ~ "Niwot Ridge Tundra",
           substr(sample_id, 1 ,1) == "c" ~ "Gordon Gulch Conifer Forest",
           substr(sample_id, 1 ,1) == "m" ~ "Gordon Gulch Meadow",
           substr(sample_id, 1 ,1) == "g" ~ "Marshall Mesa Grassland"))
# 
# inc_time_zero <- peaks_mapped %>% 
#   filter(inc_time_d == 0) %>% 
#   select(sample_id, compound, at2H, t_series_id) %>% 
#   mutate(f_start = at2H) %>% 
#   select(-at2H)
# 
#

@@

## Extract zero timepoint values
```{r}
zero_peaks <- peaks_mapped %>%
  filter(inc_time_d_str == "0 Days") %>%
  group_by(t_series_id, compound) %>%
  mutate(n_analyical_reps = n()) %>%
  ungroup() %>%
  # exclude ambiguous peaks and those with multiple data points
  # this needs to be fixed
  filter(!is_ambiguous, n_analyical_reps == 1)

zero_peaks_averaged <- zero_peaks %>% 
  group_by(soil_id, compound) %>% 
  summarise(at2H_mn_zero = mean(at2H),
            d2H_mn_zero = mean(d2H),
            area2_mn_zero = mean(area2))

zero_peaks_averaged %>% 
  openxlsx::write.xlsx(file.path("data", "IRMS", "zero_peaks_averaged.xlsx"),
                       overwrite = TRUE)

# This is the dataset we will use for generation time calculations
peaks_mapped_with_zeros <- 
  peaks_mapped %>%
  left_join(
    zero_peaks_averaged,
    by = c("soil_id", "compound")
  ) 
stopifnot(nrow(peaks_mapped_with_zeros) == nrow(peaks_mapped))

@@

## Remove standards from gencalcs
```{r}
# Don't want our standard compounds to be used for generation time calculations! x_x
peaks_mapped_with_zeros_analytes <- peaks_mapped_with_zeros %>% 
  filter(str_detect(compound, "STD", negate = TRUE))
```

@@

```{r}
gen_calc_data <- peaks_mapped_with_zeros_analytes %>% 
  mutate(
    f_label = 0.5, # label is 0.5 at%
    a = 0.5, # assimilation efficiency
    t0 = 0, # start time
    f_start = at2H_mn_zero,
    d_t = inc_time_d - t0, # change in time
    f_t = at2H, # restate 2F at time t
    u_d = gencalc(a = 0.5, # Assimilation efficiency
                  FT = f_t, # 2F at time t (at%)
                  F0 = f_start, # 2F at t0 (at%)
                  FL = f_label, # 2F of label (at%)
                  t = d_t), # change in time
         # Returns turnover (µ) in days^-1
         gen_d = log(2) / u_d # calculate generation time
  )

gen_calc_data_abbr <- gen_calc_data %>% 
  select(Analysis, sample_id, soil_id, terrain, replicate_id, compound, peak_nr, area2, 
         at2H, at2H_mn_zero,u_d, gen_d)
```

```{r}
peaks_mapped_with_zeros_analytes %>% ggplot(
  aes(x = inc_time_d, y = at2H, color = compound, group = compound)) +
  geom_point() +
  #geom_borderline() +
  scale_color_viridis_d() +
  theme_minimal()
```

@@

```{r "shaded chromatogram", fig.width=8, fig.height=8}
p_peaks_mapped_active <- peaks_mapped_with_zeros_analytes %>% ggplot(
  aes(x = rt, y = amp2, fill = as.numeric(at2H), label = compound)) +
  geom_col(aes(width = rt_end - rt_start)) +
  scale_fill_viridis_c(option = "magma") +
  facet_wrap(~sample_id, ncol = 1) +
  geom_text_repel(size = 2, box.padding = 0.5, nudge_y = 150) +
  theme_classic() +
  coord_cartesian(ylim = c(0, 1300)) +
  labs(fill = TeX("$^2F$"),
         x = "Retention Time (s)",
         y = "Amplitude (mV)") +
  theme(strip.background = element_blank())

@@

# Plots

## Generation Time of All Points

```{r}
p_gen_time_dist <- gen_calc_data %>% filter(inc_time_d != 0) %>% 
  ggplot(
  aes(x = gen_d, 
      y = inc_time_d_str, 
      color = inc_time_d_str, 
      fill = inc_time_d_str)) +
  stat_slab(height = 0.3, alpha = 0.7) +
  # stat_pointinterval(color = "black") +
  geom_boxplot(fill = "white",
               color = "black",
               width = 0.1,
               outlier.shape = NA) +
  geom_point(
    color = "black",
    #shape = 124,
    #size = 3,
    alpha = 0.5,
    position = position_nudge(y = -.03)) +
  # geom_text_repel(aes(label = compound), 
  #                 color = "black", 
  #                 nudge_y = -0.3, 
  #                 size = 3) +
  scale_x_log10() +
  annotation_logticks(sides = "b") +
  scale_fill_npg() +
  scale_color_npg() +
  theme_classic() +
  labs(x = "Generation time (days)",
       y = "Incubation time",
       title = "Generation times of soil bacteria",
       subtitle = "Lipid-specific generation times of bacteria collected from Alpine Tundra",
       caption = "Soils collected from Colorado") +
  theme(legend.position = "none",
        axis.ticks.y = element_blank())
p_gen_time_dist

@@

## Generation Time, Broken by Terrain

```{r}
p_gen_time_terrain <- gen_calc_data %>% filter(inc_time_d != 0) %>% 
  ggplot(
  aes(x = gen_d, 
      y = terrain, 
      color = terrain, 
      fill = terrain)) +
  stat_slab(height = 0.3, alpha = 0.7) +
  stat_pointinterval(color = "black") +
  geom_point(
    color = "black",
    shape = 124,
    size = 3,
    alpha = 0.5,
    position = position_nudge(y = -.03)) +
  # geom_text_repel(aes(label = compound), 
  #                 color = "black", 
  #                 nudge_y = -0.3, 
  #                 size = 3) +
  scale_x_log10() +
  annotation_logticks(sides = "b") +
  scale_fill_npg() +
  scale_color_npg() +
  theme_classic() +
  labs(x = "Generation time (days)",
       y = "Terrain",
       title = "Generation times of soil bacteria across terrain")
  theme(legend.position = "none",
        axis.ticks.y = element_blank())
p_gen_time_terrain
```

## Turnover time by compound

```{r}
gen_calc_data %>% filter(inc_time_d != 0) %>% 
  ggplot(aes(x = compound, y = u_d, color = terrain)) +
  facet_wrap(vars(terrain, inc_time_d_str)) +
  geom_point() +
  theme_bw() +
  labs(x = "Compound",
       y = "Turnover time (d^-1)",
       title = "Compound-specific turnover rates of bacterial phospholipid fatty acids") +
  theme(
    axis.text.x = element_text(angle = 90)
  )
```

## Generation Time, Broken by Terrain and Incubation Time

```{r}
p_gen_time_terrain_inc <- gen_calc_data %>% filter(inc_time_d != 0) %>% 
  ggplot(
  aes(x = gen_d, 
      y = inc_time_d_str, 
      color = soil_id, 
      fill = soil_id)) +
  stat_slab(height = 0.3, alpha = 0.7) +
  stat_pointinterval(color = "black") +
  geom_point(
    color = "black",
    shape = 124,
    size = 3,
    alpha = 0.5,
    position = position_nudge(y = -.03)) +
  # geom_text_repel(aes(label = compound), 
  #                 color = "black", 
  #                 nudge_y = -0.3, 
  #                 size = 3) +
  facet_wrap(~soil_id, ncol = 1) +
  scale_x_log10() +
  annotation_logticks(sides = "b") +
  scale_fill_npg() +
  scale_color_npg() +
  theme_classic() +
  labs(x = "Generation time (days)",
       y = "Terrain",
       title = "Generation times of soil bacteria across terrain")
  theme(legend.position = "none",
        axis.ticks.y = element_blank())
p_gen_time_terrain_inc
```

## Turnover Time, Broken By Terrain

```{r}
p_turnover_terrain <- gen_calc_data %>% filter(inc_time_d != 0) %>% 
  ggplot(
  aes(x = inc_time_d_str, 
      y = u_d, 
      color = terrain, 
      fill = terrain)) +
  #stat_slab(height = 0.3, alpha = 0.7) +
  stat_pointinterval(color = "black") +
  geom_point(
    color = "black",
    #shape = 124,
    size = 1,
    alpha = 0.5,
    position = position_nudge(x = -.03)) +
  # geom_text_repel(aes(label = compound), 
  #                 color = "black", 
  #                 nudge_y = -0.3, 
  #                 size = 3) +
  facet_wrap(~terrain, nrow = 1) +
  scale_y_log10() +
  annotation_logticks(sides = "a") +
  scale_fill_npg() +
  scale_color_npg() +
  theme_bw() +
  labs(x = "Generation time (days)",
       y = "Terrain",
       title = "Generation times of soil bacteria across terrain")
  theme(legend.position = "none",
        axis.ticks.y = element_blank())
p_turnover_terrain
```

## 7 Day Turnover Times By Terrain

```{r}
p_turnover_terrain_7d <- gen_calc_data %>% filter(inc_time_d == 7) %>% 
  ggplot(
  aes(x = terrain, 
      y = u_d)) +
  #stat_slab(height = 0.3, alpha = 0.7) +
  #stat_pointinterval(color = "black") +
  geom_boxplot(outlier.shape = NA) +
  geom_point(
    color = "black",
    #shape = 124,
    size = 1,
    alpha = 0.5) +
  # geom_text_repel(aes(label = compound), 
  #                 color = "black", 
  #                 nudge_y = -0.3, 
  #                 size = 3) +
  scale_y_log10() +
  annotation_logticks(sides = "a") +
  scale_fill_npg() +
  scale_color_npg() +
  theme_bw() +
  labs(x = "Terrain",
       y = "Turnover time (d^-1)",
       title = "Generation times of soil bacteria across terrain") +
  theme(legend.position = "none",
        axis.ticks.y = element_blank())
p_turnover_terrain_7d
```

## 7 Day Generation Times across terrain

```{r}
p_gen_time_terrain_7d <- gen_calc_data %>% filter(inc_time_d == 7) %>% 
  ggplot(
  aes(x = terrain, 
      y = gen_d)) +
  #stat_slab(height = 0.3, alpha = 0.7) +
  #stat_pointinterval(color = "black") +
  geom_boxplot(outlier.shape = NA) +
  geom_point(
    color = "black",
    #shape = 124,
    size = 1,
    alpha = 0.5) +
  # geom_text_repel(aes(label = compound), 
  #                 color = "black", 
  #                 nudge_y = -0.3, 
  #                 size = 3) +
  scale_y_log10() +
  annotation_logticks(sides = "a") +
  scale_fill_npg() +
  scale_color_npg() +
  theme_bw() +
  labs(x = "Terrain",
       y = "Turnover time (d^-1)",
       title = "Generation times of soil bacteria across terrain") +
  theme(legend.position = "none",
        axis.ticks.y = element_blank())
p_gen_time_terrain_7d
```

## Literature Rates of Cell Turnover
